---
title: "00 · Check your setup"
output:
  html_document:
    toc: false
---

**Run this once, before the workshop.** It teaches nothing — it only proves your
environment works, end to end: the packages load, your folders are connected, and a real
FCS file can be opened and plotted. Better to find a broken laptop now than at half past
eight on the morning.

**To run it:** click **Knit** in the toolbar just above this text (`Ctrl+Shift+K` on
Windows, `Cmd+Shift+K` on a Mac). A report opens in a new pane; the answer is its last line.

If you would rather watch it happen step by step, use **Run ▸ Run All** instead
(`Ctrl+Alt+R` / `Cmd+Option+R`) — the Run menu is at the top right of this editor pane, and
the output appears underneath each block of code.

Either way, look for a line starting `SETUP OK` and reply to the workshop email with it.

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## 1. R and Bioconductor versions

```{r versions}
cat("R version:      ", R.version.string, "\n", sep = "")
cat("Platform:       ", R.version$platform, "\n", sep = "")
cat("Bioconductor:   ", as.character(BiocManager::version()), "\n", sep = "")
```

## 2. Every package loads

```{r packages}
# The packages the workshop depends on. If any of these fail to load, the image is
# either the wrong one or was not pulled completely.
needed <- c("flowCore", "flowWorkspace", "openCyto", "ggcyto", "CytoML",
            "flowAI", "PeacoQC",
            "CATALYST", "FlowSOM", "uwot",
            "speedyflowplot")

missing <- character(0)

for (this_package in needed) {

  # require() tries to load a package and returns TRUE or FALSE instead of stopping.
  # suppressPackageStartupMessages() hides the banner some packages print on load.
  loaded <- suppressPackageStartupMessages(
    require(this_package, character.only = TRUE, quietly = TRUE)
  )

  if (loaded) {
    cat("ok      ", this_package, "\n", sep = "")
  } else {
    cat("MISSING ", this_package, "\n", sep = "")
    missing <- c(missing, this_package)
  }
}
```

## 3. The folders are connected

```{r folders}
# These two paths are the folders from your own computer, appearing inside the container.
data_dir     <- "/home/rstudio/data"
workshop_dir <- "/home/rstudio/workshop"

cat("data folder exists:     ", dir.exists(data_dir), "\n", sep = "")
cat("workshop folder exists: ", dir.exists(workshop_dir), "\n", sep = "")

# The workshop folder has to be writable, because that is where your work gets saved.
test_file <- file.path(workshop_dir, ".write-test")

write_worked <- tryCatch({
  writeLines("test", test_file)
  file.remove(test_file)
  TRUE
}, error = function(e) FALSE)

cat("workshop folder writable:", write_worked, "\n")
```

```{r data-files}
# There should be at least one FCS file -- the quality-control sample sent with the setup
# bundle. The rest arrive on the day.
fcs_found <- list.files(data_dir, pattern = "\\.fcs$", ignore.case = TRUE)

cat("FCS files visible: ", length(fcs_found), "\n", sep = "")

if (length(fcs_found) > 0) {
  cat(paste("  ", fcs_found, collapse = "\n"), "\n")
}
```

## 4. Read a file and draw a plot

The real test: open the quality-control file and plot it. If this works, everything the
workshop needs is working.

```{r read-and-plot, fig.width = 5.5, fig.height = 5.5}
# scales has to be attached for speedyflowplot's axis labels. Everything else was
# already attached by the loop in section 2.
library(scales)

plot_worked <- FALSE

if (length(fcs_found) > 0) {

  # Open the first FCS file in the data folder.
  #   transformation = FALSE      keep the raw recorded values
  #   truncate_max_range = FALSE  keep events above the stated maximum instead of
  #                               clamping them, which would alter the brightest cells
  first_fcs <- file.path(data_dir, fcs_found[1])

  ff <- read.FCS(first_fcs, transformation = FALSE, truncate_max_range = FALSE)

  cat("File:      ", basename(first_fcs), "\n", sep = "")
  cat("Events:    ", nrow(ff), "\n", sep = "")
  cat("Parameters:", ncol(ff), "\n", sep = "")

  # FSC-A is cell size, SSC-A is internal complexity. Every panel has them, and together
  # they are the plot every cytometry analysis starts from.
  channel_names <- colnames(ff)
  have_scatter  <- ("FSC-A" %in% channel_names) && ("SSC-A" %in% channel_names)

  if (have_scatter) {

    # PlotFlowFrame() comes from speedyflowplot. It draws a density dot plot quickly,
    # even with hundreds of thousands of events.
    PlotFlowFrame(
      ff,
      channels     = c("FSC-A", "SSC-A"),
      title        = "FSC-A vs SSC-A",
      shownumbers  = TRUE,
      plot.margins = c(3.5, 3.5, 2, 1)
    )

    plot_worked <- TRUE

  } else {
    cat("\nThis file has no FSC-A / SSC-A channels, so the plot was skipped.\n")
    cat("That is unusual but not necessarily a problem.\n")
    plot_worked <- TRUE
  }
}
```

You should see a cloud of dots. Cell size increases left to right, and granularity bottom
to top — so lymphocytes sit low and left, and granulocytes up and to the right. You are not
expected to interpret it yet; it only has to appear.

## 5. Verdict

```{r verdict}
everything_ok <- (length(missing) == 0) && dir.exists(data_dir) &&
                 write_worked && (length(fcs_found) > 0) && plot_worked

if (everything_ok) {

  # getRversion() and BiocManager::version() both return version OBJECTS, not text.
  # cat() can only print plain values, so wrap each in as.character() first.
  r_version    <- as.character(getRversion())
  bioc_version <- as.character(BiocManager::version())

  cat("SETUP OK  -  R", r_version,
      "| Bioc", bioc_version,
      "|", R.version$platform, "\n")

} else {
  cat("SETUP INCOMPLETE\n")
  if (length(missing) > 0) {
    cat("Packages that did not load:", paste(missing, collapse = ", "), "\n")
  }
  if (!dir.exists(data_dir)) {
    cat("The data folder is not connected. Did you start with `docker compose up -d`\n")
    cat("from inside the workshop folder?\n")
  }
  if (!write_worked) {
    cat("The workshop folder is not writable.\n")
  }
  if (length(fcs_found) == 0) {
    cat("No FCS file found. The setup bundle includes one -- check that the data folder\n")
    cat("from the zip is sitting next to docker-compose.yml.\n")
  }
  if (!plot_worked) {
    cat("The FCS file could not be read or plotted.\n")
  }
  cat("\nReply to the workshop email with this output and we will sort it out.\n")
}
```
