---
title: "01 · Import, compensate, transform and plot an FCS file"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
params:
  fcs_path: "/home/rstudio/data/QC251006 PB NCC_C_tubeT_C12.fcs"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 6, fig.height = 5)
```

## 1. Setup

```{r packages}
library(flowCore)      # read FCS, compensation, transformations
library(flowWorkspace) # flowjo_* transforms, GatingSet (later sessions)
library(ggcyto)        # ggplot2-based cytometry plots
library(tidyverse)
library(patchwork)     # combine plots side by side

# ggcyto facets by sample name by default; with a single file that strip is noise.
no_strip <- theme(strip.background = element_blank(), strip.text = element_blank())
```

## 2. Import

Change `fcs_path` in the header (or knit with parameters) to use a different file.
Files in the `data/` folder on your laptop appear at `/home/rstudio/data`.

```{r import}
fcs_path <- params$fcs_path
if (!file.exists(fcs_path)) {
  stop("FCS file not found: ", fcs_path,
       "\nPut your .fcs file in the ./data folder next to docker-compose.yml.")
}

# transformation = FALSE keeps raw values; truncate_max_range = FALSE keeps
# negative/over-range values that are common after digital acquisition.
ff <- read.FCS(fcs_path, transformation = FALSE, truncate_max_range = FALSE)
ff
```

```{r dims}
cat("Events:", nrow(ff), "\nParameters:", ncol(ff), "\n")
```

## 3. Metadata

### Channels and markers

Each parameter has a **channel name** (`name`, the detector, e.g. `UV379-A`) and an
optional **marker description** (`desc`, e.g. `CD45_BUV395`).

```{r parameters}
channels <- pData(parameters(ff)) |>
  as_tibble() |>
  select(name, desc, range, minRange, maxRange)
channels
```

### Keywords

The FCS TEXT segment stores acquisition metadata as key/value pairs.

```{r keywords}
kw <- keyword(ff)
cat("Total keywords:", length(kw), "\n")

selected <- c("$FIL", "$CYT", "CYTOMETER CONFIG NAME", "CREATOR", "$DATE",
              "$BTIM", "$ETIM", "$TOT", "$PAR", "TUBE NAME", "EXPERIMENT NAME")
tibble(keyword = selected, value = unlist(kw[selected]))
```

## 4. Compensation

The acquisition software stores the spillover matrix in the file (`$SPILL` or `SPILL`).

```{r spillover}
spill <- spillover(ff)$SPILL
dim(spill)
```

```{r spill-heatmap, fig.width = 9, fig.height = 8}
spill |>
  as_tibble(rownames = "from") |>
  pivot_longer(-from, names_to = "to", values_to = "spill") |>
  mutate(from = factor(from, levels = colnames(spill)),
         to   = factor(to,   levels = rev(colnames(spill)))) |>
  ggplot(aes(from, to, fill = spill)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "firebrick", limits = c(0, NA)) +
  labs(x = "Spilling from", y = "Spilling into", fill = "Spillover") +
  theme_minimal(base_size = 8) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

Find the largest spillover between two channels that both carry a marker
(unused detectors are ignored), then apply compensation:

```{r compensate}
labelled <- !is.na(channels$desc[match(colnames(spill), channels$name)])
off_diag <- spill
diag(off_diag) <- 0
off_diag[!labelled, ] <- 0
off_diag[, !labelled] <- 0
idx <- which(off_diag == max(off_diag), arr.ind = TRUE)[1, ]
from_ch <- colnames(spill)[idx[1]]
to_ch   <- colnames(spill)[idx[2]]
cat("Largest spillover:", from_ch, "->", to_ch, "=", round(max(off_diag), 3), "\n")

ff_comp <- compensate(ff, spill)
```

To compare before and after, temporarily apply the same logicle scale to both:

```{r comp-before-after, fig.width = 10, fig.height = 5}
vis_trans <- estimateLogicle(ff_comp, channels = c(from_ch, to_ch))

pair_plot <- function(x, title) {
  ggcyto(transform(x, vis_trans), aes(x = !!sym(from_ch), y = !!sym(to_ch))) +
    geom_hex(bins = 128) +
    ggtitle(title) + no_strip
}
as.ggplot(pair_plot(ff, "Uncompensated")) + as.ggplot(pair_plot(ff_comp, "Compensated"))
```

## 5. Transformation

Fluorescence data span several decades and include negative values after compensation,
so we apply a **logicle** (biexponential) transform to fluorescence channels only.
Scatter (`FSC`, `SSC`) and `Time` stay linear.

```{r transform}
fl_channels <- colnames(spill)
trans <- estimateLogicle(ff_comp, channels = fl_channels)
ff_trans <- transform(ff_comp, trans)
```

Helper to look up a channel by marker name (so the code works with other panels):

```{r marker-lookup}
marker_channel <- function(ff, marker) {
  p <- pData(parameters(ff))
  hit <- p$name[which(str_detect(p$desc, fixed(marker)))]
  if (length(hit) == 0) stop("Marker not found: ", marker)
  hit[1]
}
cd45 <- marker_channel(ff, "CD45_")
cd45
```

```{r transform-before-after, fig.width = 10, fig.height = 4}
# A handful of extreme events stretch the axes. `trim()` keeps events between the
# given quantiles of each channel using a rectangleGate -- your first gate!
trim <- function(x, chs, probs = c(0.001, 0.999)) {
  bounds <- lapply(set_names(chs), \(ch) quantile(exprs(x)[, ch], probs))
  Subset(x, rectangleGate(filterId = "on-scale", .gate = bounds))
}

hist_plot <- function(x, title) {
  ggcyto(trim(x, cd45), aes(x = !!sym(cd45))) + geom_histogram(bins = 200) +
    ggtitle(title) + no_strip
}
as.ggplot(hist_plot(ff_comp, "CD45 — linear")) +
  as.ggplot(hist_plot(ff_trans, "CD45 — logicle"))
```

## 6. Plots

```{r plots, fig.width = 10, fig.height = 9}
# Linear scatter channels have a few extreme events (clogs, coincidence) that
# squash the plot, so trim FSC/SSC to the 99.9th percentile for display.
ff_plot <- trim(ff_trans, c("FSC-A", "SSC-A"), probs = c(0, 0.999))
cat("Events shown:", nrow(ff_plot), "of", nrow(ff_trans), "\n")

dot <- function(x_ch, y_ch, title) {
  as.ggplot(
    ggcyto(ff_plot, aes(x = !!sym(x_ch), y = !!sym(y_ch))) +
      geom_hex(bins = 128) +
      ggtitle(title) + no_strip +
      theme(legend.position = "none")
  )
}

cd3  <- marker_channel(ff, "CD3_")
cd19 <- marker_channel(ff, "CD19_")
cd4  <- marker_channel(ff, "CD4_")
cd8  <- marker_channel(ff, "CD8_")

(dot("FSC-A", "SSC-A", "FSC-A vs SSC-A") | dot(cd45, "SSC-A", "CD45 vs SSC-A")) /
  (dot(cd3, cd19, "CD3 vs CD19") | dot(cd4, cd8, "CD4 vs CD8"))
```

`ggcyto` labels axes with `channel marker` automatically. Note these plots show
**all events** — gating to lymphocytes/singlets comes in a later session.

## 7. Session info

```{r session}
sessionInfo()
```
